---
title: "HW 04"
author: "Weston Scott"
format: 
  html:
    embed-resources: true
    fig-alt: true
toc: true
editor:
  render-on-save: true
execute:
  warning: false
bibliography: references.bib
---

```{r library_setup, include=FALSE}
if (!require("pacman")) 
    install.packages("pacman")

# Use pacman::p_load to install and load CRAN packages
pacman::p_load(
    dplyr,
    ggplot,
    ggrepel,
    patchwork,
    RColorBrewer,
    readxl,
    scales,
    sf,
    tigris,
    tidyverse
)

# Handle GitHub package separately
if (!require("dsbox")) {
      # Install devtools if not present
      if (!require("devtools")) 
          install.packages("devtools")
devtools::install_github("tidyverse/dsbox")
      library(dsbox)
}

options(tigris_use_cache = TRUE)

ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))
options(width = 65)

knitr::opts_chunk$set(
  fig.width = 7.5,
  fig.asp = 0.718,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)
```

## 1 - A second chance

```{r second_chance_problem1}

```

## 2. Arizona state of counties

```{r az_counties_problem2}
az_counties <- counties(state = "AZ",
                        year = 2021,
                        progress_bar = FALSE) |>
    mutate(
        name = gsub("\\s+County$", "", NAME),
        x = st_coordinates(st_centroid(geometry))[, 1],
        y = st_coordinates(st_centroid(geometry))[, 2]
    ) |> glimpse()


ggplot(az_counties) +
geom_sf(fill = 'grey90', 
        color = "grey10") +

geom_label_repel(aes(x = x, 
                     y = y, 
                     label = NAME),
                 size = 3,
                 min.segment.length = 0.1,
                 box.padding = 0.5,
                 segment.color = "grey20") +

coord_sf() +
labs(
    title = "Counties in Arizona State",
    caption = "Source: Shapefile obtained using {tigris} R package, v2.2.1",
    x = "Longitude",
    y = "Latitude"
)
```

### Resources

Found help with the labeling using the `st_coordinates` and the `geom_label_repel` [@ggrepel_gallery],[@sf_manual].

## 3. Arizona state of population change

```{r}
az_counties <- counties(
    state = "AZ",
    year = 2021,
    progress_bar = FALSE
)

pop_data <- read_excel("data/co-est2023-pop-04.xlsx", 
                       skip = 5, 
                       n_max = 15,
                       col_names = c("county", "base_2020", "pop_2020", 
                                     "pop_2021", "pop_2022", "pop_2023"))

pop_data <- pop_data |>
    mutate(
        county = sub("\\.(.+) County, Arizona", "\\1", county),
        total_pop_change_20_23 = pop_2023 - pop_2020
    ) |>
    select(-c(base_2020, pop_2020, pop_2021, pop_2022, pop_2023))

pop_data

az_data <- az_counties |>
    left_join(
        pop_data, 
        by = c("NAME" = "county")
    )

rdbu_palette <- rev(brewer.pal(5, "RdBu"))

ggplot(data = az_data) +
geom_sf(aes(fill = total_pop_change_20_23), 
        color = "white") +

scale_fill_gradientn(colors = rdbu_palette,
                    name = "Population change",
                    labels = function(x) format(x, big.mark = ",")) +

coord_sf() +
labs(
    title = "Resident Population Change for Counties in AZ",
    subtitle = "July 01, 2020 to July 01, 2023",
    caption = "Source: Shapefile obtained using {tigris} R package, v2.2.1\npopulation change data from the US Census Bureau",
    x = "Longitude",
    y = "Latitude"
) +

theme(
    plot.title.position = "plot"
)
```

### Resources

Found help with the color filling using the `scale_fill_gradientn` function and then the `rdbu` palette as well [@ggplot2_scale_gradient], [@rcolorbrewer_gallery].

## 4. Arizona state of Indiginous Tribal Regions

```{r az_tribal_regions_problem4}
az_counties <- counties(
    state = "AZ",
    year = 2021,
    progress_bar = FALSE
)

tribal_data <- st_read("data/American_Indian_Reservations_in_Arizona.shp") |>
    st_transform(crs = st_crs("EPSG:4269")) |>
    mutate(
        x = st_coordinates(st_centroid(geometry))[, 1],
        y = st_coordinates(st_centroid(geometry))[, 2]
    ) |>
    glimpse()

ggplot(az_counties) +
geom_sf(
    fill = 'grey90', 
    color = "white"
) +

geom_sf(
    data = tribal_data, 
    linewidth = 1, 
    fill = NA, 
    color = "black"
) +

geom_label_repel(
    data = tribal_data |>     
        filter(NAME %in% c("Hopi Tribe", 
                           "Navajo Nation", 
                           "White Mountain Apache Tribe", 
                           "San Carlos Apache Tribe", 
                           "Tohono O’odham Nation")),
    aes(x = x, 
        y = y, 
        label = NAME),
    size = 4,
    min.segment.length = 0.1,
    box.padding = 0.5,
    segment.color = "grey20"
) +

coord_sf() +
labs(
    title = "Indigenous Tribal Boundaries in AZ",
    caption = "Source: Shapefile obtained using {tigris} R package, v2.2.1\nIndigenous Tribe Shapefile obtained from AZGeo Data",
    x = "Longitude",
    y = "Latitude"
) +

theme(
    plot.title.position = "plot"
)
```

### Resources

Found help with reading a shapefile using the `st_read` function [@rgraphgallery_shapefile]. Found help for using `st_transform` for converting coordinate systems [@heiss_dataviz_example].

## 5. Arizona state of patchwork

```{r patchwork_plot_problem5}
az_counties <- counties(state = "AZ",
                        year = 2021,
                        progress_bar = FALSE) |>
    mutate(
        name = gsub("\\s+County$", "", NAME),
        x = st_coordinates(st_centroid(geometry))[, 1],
        y = st_coordinates(st_centroid(geometry))[, 2]
    )

tribal_data <- st_read("data/American_Indian_Reservations_in_Arizona.shp") |>
    st_transform(crs = st_crs("EPSG:4269")) |>
    mutate(
        x = st_coordinates(st_centroid(geometry))[, 1],
        y = st_coordinates(st_centroid(geometry))[, 2]
    )

pop_data <- read_excel("data/co-est2023-pop-04.xlsx", 
                       skip = 5, 
                       n_max = 15,
                       col_names = c("county", "base_2020", "pop_2020", 
                                     "pop_2021", "pop_2022", "pop_2023"))

pop_data <- pop_data |>
    mutate(
        county = sub("\\.(.+) County, Arizona", "\\1", county),
        total_pop_change_20_23 = pop_2023 - pop_2020
    ) |>
    select(-c(base_2020, pop_2020, pop_2021, pop_2022, pop_2023))

az_data <- az_counties |>
    left_join(
        pop_data, 
        by = c("NAME" = "county")
    )

rdbu_palette <- rev(brewer.pal(5, "RdBu"))

main_plot <- ggplot(data = az_data) +

    geom_sf(aes(fill = total_pop_change_20_23), 
            color = "white") +
    
    scale_fill_gradientn(colors = rdbu_palette,
                         name = "Population change",
                         labels = function(x) format(x, big.mark = ","),
                         guide = guide_colorbar(barwidth = 10,
                                                barheight = 1,
                                                direction = "horizontal",
                                                title.position = "top")
                        ) +

    geom_label_repel(data = filter(az_counties, name %in% c("Maricopa", "Pinal", "Pima")),
                     aes(x = x, 
                         y = y, 
                         label = NAME),
                     size = 4,
                     min.segment.length = 0.1,
                     box.padding = 0.5,
                     segment.color = "grey20") +

    coord_sf(xlim = c(-121, -109), ylim = c(28.5, 37)) +
    labs(
        title = NULL,
        subtitle = NULL,
        caption = NULL,
        x = NULL,
        y = NULL
    ) +
    
    theme(
        legend.position = c(0.05, 0.75),  # Adjust as needed
        legend.justification = c(0, 0),
    ) #+

    # geom_segment(data = data.frame(
    #     x = c(-116, -112),  # Corners of zoom area
    #     y = c(32.0, 32.0)
    #     # xend = c(-114, -109, -109, -114),
    #     # yend = c(32, 32, 36, 36)
    # ),
    # aes(x = x, y = y,),
    # color = "black",
    # size = 0.5,
    # linetype = "dashed")

zoom_plot <- ggplot() +
  geom_sf(data = filter(az_data, name %in% c("Maricopa", "Pinal", "Pima", "Santa Cruz", "Gila", "Yavapai")), aes(fill = total_pop_change_20_23), color = "white") +
  geom_sf(data = tribal_data, fill = NA, color = "black", linewidth = 0.75) +

scale_fill_gradientn(
  colors = rdbu_palette,
  name = NULL,
  labels = NULL,
  limits = range(az_data$total_pop_change_20_23)
) +
  geom_label_repel(
    data = filter(tribal_data, NAME %in% c("White Mountain Apache Tribe",
                                          "San Carlos Apache Tribe",
                                          "Tohono O’odham Nation")),
    aes(x = x, y = y, label = NAME),
    size = 3, box.padding = 0.5, min.segment.length = 0
  ) +
  coord_sf(xlim = c(-113.5, -110), ylim = c(31.25, 34.25)) +
  theme_void() +
  theme(panel.background = element_rect(fill = "grey50"),
       legend.position = "none")

final_map <- main_plot +
  inset_element(zoom_plot, left = 0.05, bottom = 0.05, right = 0.45, top = 0.45)

final_map
```